Arrows for Parallel Computations

### 2017-07-25 paper/content/benchmarks.tex
minor

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%.  However, when we increase the number of
requested cores to be 32---\ie when we use hyperthreading on 16 real
cores---the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyperthreaded cores and only 13.55 with PArrows/Multicore
Haskell. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the ``share nothing'' SMP build.

In the distributed memory setting same effect ensues. We obtain
plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
213.172 for 256 real cores. Apparently, hyperthreading in the Glasgow
grid fails to execute two parallel Haskell processes with full-fledged
parallelism. For this reason, we did not regard hyperthreaded cores in
our speedup plots in Figs.~\ref{fig:bench-rm-sm}--XXXXXXXX.

@@

The error bars ends were computed from pointwise maximum of both standard
deviations from both measurements for PArrows and non-PArrows
versions. These are the values provided by |Criterion| package that we
used for benchmarking. We call a difference between two versions
significant when the border of the error bar of absolute time
difference is above or below zero. In other words: the time
difference is significant if it is above measurement error.

>>>>

The PArrows version of \rmtest on a single node of the Glasgow grid
showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65, the
efficiency was 85.3\%.  However, if we increase the number of
requested cores to be 32---\ie if we use hyperthreading on 16 real
cores---the speedup does not increase that well. It is merely 15.99
for 32 tasks with PArrows/Eden. It is worse for other backends.  As
for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
hyperthreaded cores and only 13.55 with PArrows/Multicore
Haskell. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
version used here was Eden~CP, the ``share nothing'' SMP build.

In the distributed memory setting the same effect ensues. We obtain
plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
213.172 for 256 real cores. Apparently, hyperthreading in the Glasgow
grid fails to execute two parallel Haskell processes with full-fledged
parallelism. For this reason, we did not regard hyperthreaded cores in
our speedup plots in Figs.~\ref{fig:bench-rm-sm}--XXXXXXXX.

@@

The error bars ends were computed from pointwise maximum of both standard
deviations from both measurements for PArrows and non-PArrows
versions. These are the values provided by the |bench| package that we
used for benchmarking. We call a difference between two versions
significant when the border of the error bar of absolute time
difference is above or below zero. In other words: the time
difference is significant if it is above measurement error.

### 2017-07-25 paper/content/motivation.tex
minor

Pure functional programming languages have been used
since longer time as a ``sand box'' for testing novel parallel
programming approaches. Reasons for this include referential
transparency and the ease of introducing new sophisticated language
concepts. Specially for Haskell (on which this paper focuses), various
approaches reached maturity. The definite parallel Haskell does not
exist, there are multiple parallel Haskells. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
We use here Arrows to represent parallel computations. This has not
been done before.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and various implementations.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner. We tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends are viable, too.

\paragraph{Contribution}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but
% why do we aim to abstract parallelism this way and what does this
% approach do better than the other parallel Haskells?
 is such a parallelism abstraction of benefit and to what extent does
 it improve existing approaches?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
\end{itemize}

>>>>

Pure functional programming languages have been used
since a longer time as a ``sand box'' for testing novel parallel
programming approaches. Reasons for this include referential
transparency and the ease of introducing new sophisticated language
concepts. Specially for Haskell (on which this paper focuses), various
approaches reached maturity. The definite parallel Haskell does not
exist, there are multiple parallel Haskells \mbcomment{Wortwiederholung? Oder okay so?}. We regard here in-depth
the Multicore Haskell (a SMP implementation of GpH language), the
|Par| Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a monad, even if
only for the internal representation. Some introduce additional
language constructs. 
%In this paper we introduce a notion of parallel
%computations using Arrows.
Here, we use Arrows to represent parallel computations. This has not
been done before.

%Our Arrows-based interface is a high-level one.
We provide an Arrows-based typeclass and various implementations.
Instead of 
introducing a new low-level parallel backend in order to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface and varying implementations in
existing parallel Haskells: Multicore Haskell, |Par| Monad, and
Eden. Thus, we not only define a parallel programming interface in a
novel manner - we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel Haskell backends at will. Further backends are viable, too.

\paragraph{Contribution}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
\mbcomment{different, how?}
We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
% why do we aim to abstract parallelism this way and what does this
% approach do better than the other parallel Haskells?
 is such a parallelism abstraction of benefit and to what extent does
 it improve existing approaches?
\begin{itemize}
	\item \textbf{Arrow DSL benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator. The resulting Arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic. This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell. So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
\end{itemize}
